---
title: "Exploring the Growth Rate of the stringy05 Index"
date: 2026-08-30
categories: [scagnostics, minimum spanning tree, stringy05, asymptotic analysis, R]
description: "A theoretical and empirical exploration of the stringy05 scagnostic index."
license: "All Rights Reserved"
copyright: "Copyright © 2026 Tina Rashid Jafari. All rights reserved."
bibliography: references.bib
format:
  html:
    toc: true
    toc-depth: 3
execute:
  echo: true
code-fold: true
warning: FALSE
message: FALSE
---

## Aim

The goal of this blog is to understand the asymptotic behavior of the `stringy05` index, defined as

$$
\text{stringy05} =
\frac{{diameter}(T_n)}{{length}(T_n)}.
$$

Here $T_n$ is the Euclidean minimum spanning tree (MST) built on $n$ random bivariate observations.

The denominator is the total Euclidean length of the MST,

$$
L_n
= {length}(T_n) = \sum_{e\in T_n}|e|.
$$

The numerator is the weighted graph diameter of the MST,

$$
D_n
= {diameter}(T_n)
= \max_{u,v\in T_n} d_{T_n}(u,v),
$$
where $d_{T_n}(u,v)$ is the length of the unique path in the tree from $u$ to $v$, using Euclidean edge lengths as weights.

Thus,

$$
R_n
=
\text{stringy05}(T_n)
=
\frac{D_n}{L_n}.
$$

The main question is whether

$$
R_n \to 0,
$$

or whether

$$
R_n \to \rho > 0.
$$

Empirically, previous simulations suggested that the ratio does not rapidly decay toward zero. Around sample sizes near $n=900$, the estimated 95th percentile of the noise distribution appeared to stabilize near $0.1$. This motivates a more detailed investigation of the numerator and denominator separately.


## Denominator: theoretical growth rate of total MST length

The denominator $L_n$ is the part of the problem for which we do have strong theoretical guidance.

@steele1988growth studies Euclidean minimal spanning trees with power-weighted edges. In his notation, for points $x_1,\dots,x_n$, the MST functional is

$$
M(x_1,\dots,x_n)
=
\min_T \sum_{e\in T}\psi(|e|),
$$

where the minimum is over spanning trees $T$, and $\psi$ is an edge-weighting function [@steele1988growth].

For the ordinary Euclidean MST length, we take

$$
\psi(t)=t.
$$

Then

$$
M(X_1,\dots,X_n)
=
\sum_{e\in T_n}|e|
=
L_n.
$$

So Steele's $M_n$ [@steele1988growth] is exactly the denominator of the stringy05 index.

## Steele's growth theorem

@steele1988growth considers weight functions satisfying

$$
\psi(t)\sim t^\alpha
\quad\text{as } t\to 0,
$$

with

$$
0<\alpha<d.
$$


His main theorem states that, for iid observations in $\mathbb R^d$ from a compactly supported distribution, if $f$ is the density of the absolutely continuous part of the distribution, then

$$
n^{-(d-\alpha)/d}
M(X_1,\dots,X_n)
\to
c(\alpha,d)
\int f(x)^{(d-\alpha)/d}\,dx
$$

almost surely.


In the ordinary MST length case,

$$
\alpha=1.
$$

For bivariate data,

$$
d=2.
$$

Therefore,

$$
n^{-1/2}L_n
\to
c(1,2)
\int f(x)^{1/2}\,dx.
$$

Equivalently,

$$
L_n
\sim
C_f\sqrt n,
$$

where

$$
C_f
=
c(1,2)
\int f(x)^{1/2}\,dx.
$$

Thus, for two-dimensional data, the denominator has theoretical growth rate

$$
\boxed{
L_n= O(\sqrt n).
}
$$

@steele1988growth first proves the MST growth-rate result in the uniform setting, where the sample points are distributed in the unit cube. The result is then extended beyond the uniform case by adding a density correction. For a more general distribution with density $f$, the limiting constant is modified by a factor of the form

$$
\int f(x)^{(d-\alpha)/d}\,dx.
$$

In the ordinary bivariate MST-length case, this becomes

$$
\int f(x)^{1/2}\,dx.
$$

This extension is important for my project because the simulations are based on bivariate noise distributions rather than only uniform samples.

However, there is one important technical caveat. The theorem in @steele1988growth is stated under compact-support assumptions. A standard bivariate normal distribution has unbounded support, so it falls outside the direct assumptions of the theorem. Therefore, for Gaussian samples, I use the $\sqrt n$ scaling from @steele1988growth as a theoretically motivated benchmark for the denominator, not as a theorem that applies directly without further work. 


## Numerator: what is known about the MST diameter?

The numerator is

$$
D_n
=
\max_{u,v}d_{T_n}(u,v).
$$

This is the weighted longest shortest path in the MST. It is a global path functional: it depends on how many edges lie along the longest path and how their lengths accumulate.

This is different from the total MST length $L_n$, which is a sum over all edges. It is also different from the longest single edge of the MST.

At this point, the exact theoretical growth rate of $D_n$ is not given by the MST length theorem in @steele1988growth.


## Penrose's longest-edge theorem and its limitation

@penrose1997longest studies the longest edge of the random Euclidean MST. Let

$$
E_n
=
\max_{e\in T_n}|e|.
$$

For uniform random points in the unit square, @penrose1997longest shows that the longest MST edge has the same asymptotic extreme-value behavior as the longest nearest-neighbor edge. In dimension $2$, the result implies that

$$
n\pi E_n^2-\log n
$$

converges in distribution to a double-exponential limiting law [@penrose1997longest].


@penrose1997longest also shows that very long MST edges usually behave like nearest-neighbor edges. In other words, the longest edges are typically attached to leaf vertices, or endpoints, of the MST.

This matters for the stringy05 index because the MST diameter is not just the longest single edge. The diameter is the length of a whole path through the tree. So the result in @penrose1997longest suggests that the longest individual edges are often at the outside of the tree, rather than forming the main interior path that determines the MST diameter.


The longest edge scale is therefore approximately

$$
E_n
\asymp
\sqrt{\frac{\log n}{n}}.
$$
However, this is not the same as the stringy05 numerator.

The stringy05 numerator is

$$
D_n =
\max_{u,v} d_{T_n}(u,v),
$$

which is the length of an entire path. This path may contain many MST edges. Therefore, the theorem in @penrose1997longest does not give the growth rate of $D_n$. It only gives the scale of the largest individual edge.


## A simple example: longest edge versus MST diameter

Let's imagine the MST is a path with five vertices:

$$
A \;-\; B \;-\; C \;-\; D \;-\; E
$$

and the edge weights are

$$
|AB|=0.10,\quad |BC|=0.12,\quad |CD|=0.09,\quad |DE|=0.11.
$$

Penrose's longest-edge quantity looks only for the single largest edge in the MST:

$$
E_n=\max_{e\in T_n}|e|=0.12.
$$

So in this example, Penrose's quantity is the edge $BC$.

But the stringy05 numerator is the weighted MST diameter. It looks for the longest path in the tree and adds all edge weights along that path. Here the longest path is from $A$) to $E$:

$$
D_n = |AB|+|BC|+|CD|+|DE| =
0.10+0.12+0.09+0.11 =
0.42.
$$

So in this example,

$$
\text{Penrose longest edge}=0.12,
$$

while

$$
\text{stringy05 diameter}=0.42.
$$


```{r}
library(cassowaryr)
library(igraph)
library(ggplot2)
library(patchwork)

set.seed(1050)

n <- 200
u <- runif(n, -1, 1)
polynomial <- poly(u, degree = 2)

x <- polynomial[, 1] + rnorm(n, sd = 0.03)
y <- polynomial[, 2] + rnorm(n, sd = 0.03)

ggplot(data.frame(x, y), aes(x, y)) +
  geom_point(size = 1.6) +
  coord_equal() +
  theme_bw() +
  labs(
    x = "Polynomial degree 1",
    y = "Polynomial degree 2"
  ) +
  theme(aspect.ratio = 1,
        axis.ticks = element_blank(),
        axis.text = element_blank())
```

## Construct the minimum spanning tree

```{r}
sc <- scree(
  x,
  y,
  out.rm = FALSE,
  binner = NULL
)

mst <- cassowaryr:::gen_mst(
  sc$del,
  sc$weights
)

mst_weights <- E(mst)$weight
```

## Calculate the two diameter definitions

```{r}
# Stringy: longest weighted path through the MST
diameter_path <- get_diameter(
  mst,
  directed = FALSE,
  weights = mst_weights
)

weighted_diameter <- diameter(
  mst,
  directed = FALSE,
  weights = mst_weights
)

# Penrose: longest individual MST edge
longest_edge <- E(mst)[which.max(mst_weights)]
penrose_diameter <- longest_edge$weight

# Stringy05
stringy05 <- weighted_diameter / sum(mst_weights)

cat("Penrose longest edge:", round(penrose_diameter, 5), "\n")
cat("Weighted MST diameter:", round(weighted_diameter, 5), "\n")
cat("Stringy05:", round(stringy05, 5), "\n")
```

## Draw the MST and highlight the two diameters

```{r}
# Coordinates of the Stringy diameter path
diameter_xy <- as.data.frame(
  sc$del$x[as.integer(diameter_path), ]
)

names(diameter_xy) <- c("x", "y")

# Coordinates of the longest individual edge
penrose_vertices <- ends(
  mst,
  longest_edge,
  names = FALSE
)

penrose_xy <- as.data.frame(
  sc$del$x[penrose_vertices, ]
)

names(penrose_xy) <- c("x", "y")

# Base MST plot supplied by cassowaryr
mst_plot <- draw_mst(mst, sc) +
  coord_equal() +
  theme_bw()

p_stringy <- mst_plot +
  geom_path(
    data = diameter_xy,
    aes(x, y),
    colour = "#0072B2",
    linewidth = 1.3
  ) +
  geom_point(
    data = diameter_xy[c(1, nrow(diameter_xy)), ],
    aes(x, y),
    colour = "#0072B2",
    size = 3
  ) +
  labs(
    title = "Stringy weighted diameter",
    subtitle = paste(
      "Weighted diameter =",
      round(weighted_diameter, 4)
    )
  )

p_penrose <- mst_plot +
  geom_segment(
    data = penrose_xy,
    aes(
      x = x[1],
      y = y[1],
      xend = x[2],
      yend = y[2]
    ),
    colour = "#D55E00",
    linewidth = 1.5
  ) +
  geom_point(
    data = penrose_xy,
    aes(x, y),
    colour = "#D55E00",
    size = 3
  ) +
  labs(
    title = "Penrose longest MST edge",
    subtitle = paste(
      "Longest edge =",
      round(penrose_diameter, 4)
    )
  )

p_stringy + p_penrose +
  plot_annotation(
    subtitle = paste(
      "Stringy05 =",
      round(stringy05, 4)
    )
  )
```


This shows why the result in @penrose1997longest is useful only as a weak lower bound. The longest single edge is always less than or equal to the weighted MST diameter, but it is not the same quantity:

$$
E_n\le D_n.
$$


## Upper and lower bound for the MST diameter

Even though the longest-edge theorem in @penrose1997longest does not solve the diameter problem, it provides a weak lower bound.

For every weighted tree,

$$
\boxed{
E_n\le D_n\le L_n.
}
$$

The first inequality holds because the longest edge is itself the path between its endpoints in the tree. The second inequality holds because any path in the tree uses only a subset of the edges of the tree. 


Combining the longest-edge scale from @penrose1997longest with the total-length scale from @steele1988growth gives

$$
\sqrt{\frac{\log n}{n}}
\lesssim
D_n
\le
O(\sqrt n).
$$

Dividing by $L_n$ gives the weak theoretical ratio bound

$$
\frac{\sqrt{\log n}}{n}
\lesssim
R_n
\le
1.
$$

This lower bound is mathematically valid but very weak. For example, at $n=1000$,

$$
\frac{\sqrt{\log(1000)}}{1000}
\approx
0.00263.
$$

This is much smaller than the empirical values observed in simulation, which are closer to $0.2$. Therefore, the observed stringy05 behavior cannot be explained by the longest single MST edge alone. The diameter appears to be driven by the accumulation of many edges along a long path.



## Next step: study numerator and denominator separately

The next part of this analysis will study

$$
L_n=\operatorname{length}(T_n)
$$

and

$$
D_n=\operatorname{diameter}(T_n)
$$

separately.

The plan is:

1. Generate bivariate noise samples for a grid of sample sizes $n$.
2. Record the total MST length $L_n$ .
3. Record the weighted MST diameter $D_n$.
4. Record the ratio

$$
R_n=\frac{D_n}{L_n}.
$$

For the denominator, the theorem in @steele1988growth suggests checking whether

$$
\frac{L_n}{\sqrt n}
$$

stabilizes.

For the numerator, the key empirical question is whether

$$
\frac{D_n}{\sqrt n}
$$

also stabilizes.

If both normalized quantities stabilize, then the ratio should stabilize near a positive constant:

$$
R_n
=
\frac{D_n}{L_n}
\approx
\frac{C_D\sqrt n}{C_L\sqrt n}
=
\frac{C_D}{C_L}.
$$

This would support the empirical observation that stringy05 converges to a positive number, possibly near $0.1$, rather than to zero.



## Empirical check of the denominator growth rate

To check the theoretical growth rate of the denominator, I simulated independent bivariate observations from the uniform distribution on the unit square, which matches the setting used in @steele1988growth. For each sample size, I repeated the simulation 100 times (just for testing), constructed the Euclidean minimum spanning tree, and recorded its total edge length,

$$
L_n = \sum_{e\in T_n}|e|.
$$

The simulation results were saved in separate files for different ranges of sample sizes and are combined below.

```{r}
library(readr)
library(dplyr)
library(purrr)
library(ggplot2)

csv_files <- c(
  "stringy05_uniform_50_175.csv",
  "stringy05_uniform_200_500.csv",
  "stringy05_uniform_550_800.csv",
  "stringy05_uniform_850_1000.csv",
  "stringy05_uniform_1100_2000.csv",
  "stringy05_uniform_3000_5000.csv"
)

combined_results <- map_dfr(
  csv_files,
  ~ read_csv(.x, show_col_types = FALSE)
) |>
  arrange(n, simulation_run)
```

The result in @steele1988growth implies that, for bivariate uniform data, the total MST length grows at rate

$$
L_n \sim C\sqrt n,
$$

for some positive constant $C$.

To compare the simulations with this theoretical rate, I first calculate the empirical mean MST length for each sample size. I then estimate the multiplicative constant $C$ by fitting the relationship

$$
L_n \approx C\sqrt n
$$

through the origin.

```{r}
# Calculate the mean for each sample size
length_summary <- combined_results |>
  group_by(n) |>
  summarise(
    mean_length = mean(total_mst_length),
    .groups = "drop"
  )

# Estimate C in L_n = C * sqrt(n)
length_fit <- lm(
  mean_length ~ 0 + I(sqrt(n)),
  data = length_summary
)

C_hat <- unname(coef(length_fit)[1])

C_hat

# Theoretical curve based on the fitted constant
theoretical_curve <- data.frame(
  n = seq(
    min(combined_results$n),
    max(combined_results$n),
    length.out = 500
  )
) |>
  mutate(
    theoretical_length = C_hat * sqrt(n)
  )
```

The following plot shows all individual simulation values of the total MST length. The blue line gives the empirical mean at each sample size, while the pink dashed line shows the fitted theoretical function

$$
\widehat C\sqrt n.
$$

If the denominator follows the theoretical $\sqrt n$ growth rate, the empirical mean should have approximately the same shape as this dashed curve.

```{r}
ggplot() +
  geom_point(
    data = combined_results,
    aes(
      x = n,
      y = total_mst_length,
      colour = "Simulation values"
    ),
    alpha = 0.7,
    size = 0.5
  ) +
  geom_line(
    data = length_summary,
    aes(
      x = n,
      y = mean_length,
      colour = "Empirical mean"
    ),
    linewidth = 0.5
  ) +
  geom_point(
    data = length_summary,
    aes(
      x = n,
      y = mean_length,
      colour = "Empirical mean"
    ),
    size = 1
  ) +
  geom_line(
    data = theoretical_curve,
    aes(
      x = n,
      y = theoretical_length,
      colour = "Theoretical sqrt(n)"
    ),
    linewidth = 1,
    linetype = "dashed"
  ) +
  scale_colour_manual(
    name = NULL,
    values = c(
      "Simulation values" = "black",
      "Empirical mean" = "royalblue",
      "Theoretical sqrt(n)" = "violetred"
    )
  ) +
  labs(
    x = "Sample size n",
    y = "Total MST length",
    title = "Growth of total Euclidean MST length",
    subtitle = paste0(
      "Dashed curve: ",
      round(C_hat, 4),
      " sqrt(n)"
    )
  ) +
  theme_bw() +
  theme(
    aspect.ratio = 1,
    plot.title = element_text(face = "bold"),
    legend.position = "bottom"
  )
```

A second way to check the growth rate is to remove the proposed $\sqrt n$ scaling. If

$$
L_n \sim C\sqrt n,
$$

then

$$
\frac{L_n}{\sqrt n}\to C.
$$

Therefore, if the theoretical rate is appropriate, the normalized MST length should become increasingly stable as the sample size grows.

The plot below shows the normalized value $L_n/\sqrt n$ for every simulation. The blue line gives the empirical mean for each sample size, and the pink dashed line shows the same finite-sample estimate $\widehat C$ obtained above.

```{r}
# Normalize total MST length by sqrt(n)
combined_results <- combined_results |>
  mutate(
    length_over_sqrtn = total_mst_length / sqrt(n)
  )

# Mean normalized length for each sample size
length_scaled_summary <- combined_results |>
  group_by(n) |>
  summarise(
    mean_length_over_sqrtn = mean(length_over_sqrtn),
    .groups = "drop"
  )

ggplot() +
  geom_point(
    data = combined_results,
    aes(
      x = n,
      y = length_over_sqrtn,
      colour = "Simulation values"
    ),
    alpha = 0.7,
    size = 0.5
  ) +
  geom_line(
    data = length_scaled_summary,
    aes(
      x = n,
      y = mean_length_over_sqrtn,
      colour = "Empirical mean"
    ),
    linewidth = 0.5
  ) +
  geom_point(
    data = length_scaled_summary,
    aes(
      x = n,
      y = mean_length_over_sqrtn,
      colour = "Empirical mean"
    ),
    size = 1
  ) +
  geom_hline(
    aes(
      yintercept = C_hat,
      colour = "Estimated C"
    ),
    linewidth = 1,
    linetype = "dashed"
  ) +
  scale_colour_manual(
    name = NULL,
    values = c(
      "Simulation values" = "black",
      "Empirical mean" = "royalblue",
      "Estimated C" = "violetred"
    )
  ) +
  labs(
    x = "Sample size n",
    y = expression(L[n] / sqrt(n)),
    title = expression("Convergence of " * L[n] / sqrt(n)),
    subtitle = paste0(
      "Dashed horizontal line: C = ",
      round(C_hat, 4)
    )
  ) +
  theme_bw() +
  theme(
    aspect.ratio = 1,
    plot.title = element_text(face = "bold"),
    legend.position = "bottom"
  )
```



## Empirical check of the longest MST edge

The theoretical result for the numerator is not directly available for the weighted MST diameter used by `stringy05`. However, @penrose1997longest studies the longest **single edge** of the Euclidean minimum spanning tree. Let

$$
E_n = \max_{e \in T_n} |e|.
$$

For independent uniform observations on the unit square, the longest MST edge has asymptotic scale

$$
E_n \asymp \sqrt{\frac{\log n}{n}}.
$$

This is different from the actual `stringy05` numerator, which is the weighted length of an entire path through the MST. Therefore, in the simulations I use the stored `penrose_diameter` variable to check the longest-edge result in @penrose1997longest separately.

For each sample size, the simulation was repeated 100 times. I first calculate the empirical mean longest-edge length for each value of $n$.

```{r}
penrose_summary <- combined_results |>
  group_by(n) |>
  summarise(
    mean_penrose = mean(penrose_diameter),
    .groups = "drop"
  )
```

The result in @penrose1997longest suggests a relationship of the form

$$
E_n \approx A
\sqrt{\frac{\log n}{n}},
$$

where $A$ is a positive constant.

As with the denominator analysis, I estimate the constant $A$ from the simulation values.

```{r}
penrose_fit <- lm(
  mean_penrose ~ 0 + I(sqrt(log(n) / n)),
  data = penrose_summary
)

A_hat <- unname(coef(penrose_fit)[1])

A_hat
```

The estimated value $\widehat A$ is then used to construct the fitted theoretical curve

$$
\widehat A
\sqrt{\frac{\log n}{n}}.
$$

```{r}
penrose_theoretical_curve <- data.frame(
  n = seq(
    min(combined_results$n),
    max(combined_results$n),
    length.out = 500
  )
) |>
  mutate(
    theoretical_penrose =
      A_hat * sqrt(log(n) / n)
  )
```

The following plot shows all individual simulation values of the longest MST edge. The blue line shows the empirical mean for each sample size, while the pink dashed curve shows the fitted theoretical function

$$
\widehat A
\sqrt{\frac{\log n}{n}}.
$$

If the simulations are consistent with @penrose1997longest, the empirical longest-edge values should decrease with approximately the same shape as the theoretical curve.

```{r}
ggplot() +
  geom_point(
    data = combined_results,
    aes(
      x = n,
      y = penrose_diameter,
      colour = "Simulation values"
    ),
    alpha = 0.7,
    size = 0.5
  ) +
  geom_line(
    data = penrose_summary,
    aes(
      x = n,
      y = mean_penrose,
      colour = "Empirical mean"
    ),
    linewidth = 0.5
  ) +
  geom_point(
    data = penrose_summary,
    aes(
      x = n,
      y = mean_penrose,
      colour = "Empirical mean"
    ),
    size = 1
  ) +
  geom_line(
    data = penrose_theoretical_curve,
    aes(
      x = n,
      y = theoretical_penrose,
      colour = "Theoretical sqrt(log(n)/n)"
    ),
    linewidth = 1,
    linetype = "dashed"
  ) +
  scale_colour_manual(
    name = NULL,
    values = c(
      "Simulation values" = "black",
      "Empirical mean" = "royalblue",
      "Theoretical sqrt(log(n)/n)" = "violetred"
    )
  ) +
  labs(
    x = "Sample size n",
    y = "Longest MST edge",
    title = "Scaling of the longest Euclidean MST edge",
    subtitle = paste0(
      "Dashed curve: ",
      round(A_hat, 4),
      " sqrt(log(n)/n)"
    )
  ) +
  theme_bw() +
  theme(
    aspect.ratio = 1,
    plot.title = element_text(face = "bold"),
    legend.position = "bottom"
  )
```



## Does the stringy05 numerator follow the Penrose rate?

The previous analysis verified the theoretical result for the longest single MST edge. However, the numerator of `stringy05` is not the longest individual edge. It is the weighted MST diameter, which is the total Euclidean length of the longest path through the MST.


The next empirical question is whether the weighted diameter nevertheless follows a similar scaling,

$$
D_n
\approx
B\sqrt{\frac{\log n}{n}},
$$

for some constant $B$.

```{r}
stringy_diameter_summary <- combined_results |>
  group_by(n) |>
  summarise(
    mean_stringy_diameter = mean(stringy_diameter),
    .groups = "drop"
  )
```

To compare the weighted diameter with the functional form from @penrose1997longest, I fit

$$
D_n
\approx
B\sqrt{\frac{\log n}{n}}
$$

 The purpose is to test whether the **shape and rate** of the weighted diameter are consistent with the rate from @penrose1997longest.

```{r}
stringy_diameter_fit <- lm(
  stringy_diameter ~ 0 + I(sqrt(log(n) / n)),
  data = combined_results
)

B_hat <- unname(coef(stringy_diameter_fit)[1])

B_hat
```

```{r}
stringy_theoretical_curve <- data.frame(
  n = seq(
    min(combined_results$n),
    max(combined_results$n),
    length.out = 500
  )
) |>
  mutate(
    theoretical_stringy_diameter =
      B_hat * sqrt(log(n) / n)
  )
```

The following plot compares all simulated weighted MST diameters with the theoretical boundary from @penrose1997longest.

```{r}
ggplot() +
  geom_point(
    data = combined_results,
    aes(
      x = n,
      y = stringy_diameter,
      colour = "Simulation values"
    ),
    alpha = 0.7,
    size = 0.5
  ) +
  geom_line(
    data = stringy_diameter_summary,
    aes(
      x = n,
      y = mean_stringy_diameter,
      colour = "Empirical mean"
    ),
    linewidth = 0.5
  ) +
  geom_point(
    data = stringy_diameter_summary,
    aes(
      x = n,
      y = mean_stringy_diameter,
      colour = "Empirical mean"
    ),
    size = 1
  ) +
  geom_line(
    data = stringy_theoretical_curve,
    aes(
      x = n,
      y = theoretical_stringy_diameter,
      colour = "Penrose-type rate"
    ),
    linewidth = 1,
    linetype = "dashed"
  ) +
  scale_colour_manual(
    name = NULL,
    values = c(
      "Simulation values" = "black",
      "Empirical mean" = "royalblue",
      "Penrose-type rate" = "violetred"
    )
  ) +
  labs(
    x = "Sample size n",
    y = "Weighted MST diameter",
    subtitle = paste0(
      "Dashed curve: ",
      round(B_hat, 4),
      " sqrt(log(n)/n)"
    )
  ) +
  theme_bw() +
  theme(
    aspect.ratio = 1,
    plot.title = element_text(face = "bold"),
    legend.position = "bottom"
  )
```


## Addario-Berry, Broutin and Reed: an $n^{1/3}$ MST-diameter benchmark

@addarioberry2006diameter study a different random minimum-spanning-tree model. They begin with the complete graph $K_n$, assign independent and identically distributed continuous weights to all edges, and construct the minimum-weight spanning tree. Their main result is

$$
\mathbb{E}\{\operatorname{diam}(T_n)\}
=
\Theta(n^{1/3}).
$$

There are two important differences from my setting:

1. **Randomness of the edge weights.** In @addarioberry2006diameter, the edge weights of $K_n$ are i.i.d. In my Euclidean MST, the edge weights are distances between random points in $\mathbb{R}^2$, so the pairwise edge weights are geometrically constrained and are not independent.
2. **Definition of diameter.** Their diameter is the ordinary graph diameter, based on graph distance along the tree. My `stringy05` numerator is a **weighted Euclidean diameter**, where the Euclidean edge lengths are summed along the longest weighted path.

To see whether the empirical weighted diameter has anything resembling the same growth shape, I fit the benchmark

$$
D_n
\approx
C_{\mathrm{ABR}} n^{1/3},
$$

to the empirical weighted-diameter means.

```{r}

abr_fit <- lm(
  mean_stringy_diameter ~ 0 + I(n^(1/3)),
  data = stringy_diameter_summary
)

C_ABR_hat <- unname(coef(abr_fit)[1])


abr_theoretical_curve <- data.frame(
  n = seq(
    min(combined_results$n),
    max(combined_results$n),
    length.out = 500
  )
) |>
  mutate(
    abr_n13_benchmark = C_ABR_hat * n^(1/3)
  )
```

The following plot shows the same simulated weighted MST diameters used in the previous section, but now compares them with the fitted $n^{1/3}$ benchmark from @addarioberry2006diameter.

```{r}
ggplot() +
  geom_point(
    data = combined_results,
    aes(
      x = n,
      y = stringy_diameter,
      colour = "Simulation values"
    ),
    alpha = 0.7,
    size = 0.5
  ) +
  geom_line(
    data = stringy_diameter_summary,
    aes(
      x = n,
      y = mean_stringy_diameter,
      colour = "Empirical mean"
    ),
    linewidth = 0.5
  ) +
  geom_point(
    data = stringy_diameter_summary,
    aes(
      x = n,
      y = mean_stringy_diameter,
      colour = "Empirical mean"
    ),
    size = 1
  ) +
  geom_line(
    data = abr_theoretical_curve,
    aes(
      x = n,
      y = abr_n13_benchmark,
      colour = "Addario-Berry n^(1/3) benchmark"
    ),
    linewidth = 1,
    linetype = "dashed"
  ) +
  scale_colour_manual(
    name = NULL,
    values = c(
      "Simulation values" = "black",
      "Empirical mean" = "royalblue",
      "Addario-Berry n^(1/3) benchmark" = "violetred"
    )
  ) +
  labs(
    x = "Sample size n",
    y = "Weighted MST diameter",
    subtitle = paste0(
      "Dashed curve: ",
      round(C_ABR_hat, 4),
      " n^(1/3)"
    )
  ) +
  theme_bw() +
  theme(
    aspect.ratio = 1,
    plot.title = element_text(face = "bold"),
    legend.position = "bottom"
  )
```

## Empirical check of the Addario-Berry hop-diameter growth rate

@addarioberry2006diameter study the graph diameter of a random minimum
spanning tree, where distance is measured by the number of edges along a path.
They show that the expected diameter grows at rate

$$
E(H_n) = \Theta(n^{1/3}),
$$

where $H_n$ denotes the hop diameter.

Although their random-MST model differs from the Euclidean MST considered here,
the $n^{1/3}$ rate can be used as a comparison benchmark.

For each sample size, I simulated 100 independent bivariate uniform samples and
recorded the unweighted hop diameter of the resulting Euclidean MST.

```{r}

addario_files <- c(
  "addario_uniform_50_175.csv",
  "addario_uniform_200_500.csv",
  "addario_uniform_550_800.csv",
  "addario_uniform_850_1000.csv",
  "addario_uniform_1100_2000.csv",
  "addario_uniform_3000_5000.csv"
)

addario_results <- map_dfr(
  addario_files,
  ~ read_csv(.x, show_col_types = FALSE)
) |>
  arrange(n, simulation_run)
```

For each value of $n$, the empirical mean hop diameter is

$$
\overline H_n
=
\frac{1}{100}
\sum_{b=1}^{100} H_{n,b}.
$$
```{r}
addario_summary <- addario_results |>
  group_by(n) |>
  summarise(
    mean_hop_diameter = mean(addario_hop_diameter),
    .groups = "drop"
  )
```


### Fit the $n^{1/3}$ growth rate

The theorem specifies the growth rate but not an exact multiplicative constant.
Therefore, I estimate the constant $C$ in

$$
E(H_n) \approx Cn^{1/3}
$$

from the simulated mean diameters.

```{r}
addario_fit <- lm(
  mean_hop_diameter ~ 0 + I(n^(1/3)),
  data = addario_summary
)

C_addario <- unname(coef(addario_fit)[1])

C_addario
```

Then create the theoretical curve:

```{r}
addario_curve <- data.frame(
  n = seq(
    min(addario_results$n),
    max(addario_results$n),
    length.out = 500
  )
) |>
  mutate(
    theoretical_hop_diameter =
      C_addario * n^(1/3)
  )
```

## Plotting hop diameter and $n^{1/3}$ growth


```{r}
ggplot() +
  geom_point(
    data = addario_results,
    aes(
      x = n,
      y = addario_hop_diameter,
      colour = "Simulation values"
    ),
    alpha = 0.35,
    size = 0.6
  ) +
  
  geom_line(
    data = addario_summary,
    aes(
      x = n,
      y = mean_hop_diameter,
      colour = "Empirical mean"
    ),
    linewidth = 0.7
  ) +
  
  geom_point(
    data = addario_summary,
    aes(
      x = n,
      y = mean_hop_diameter,
      colour = "Empirical mean"
    ),
    size = 1.4
  ) +
  
  geom_line(
    data = addario_curve,
    aes(
      x = n,
      y = theoretical_hop_diameter,
      colour = "Fitted n^(1/3) rate"
    ),
    linewidth = 1,
    linetype = "dashed"
  ) +
  
  scale_colour_manual(
    name = NULL,
    values = c(
      "Simulation values" = "black",
      "Empirical mean" = "royalblue",
      "Fitted n^(1/3) rate" = "violetred"
    )
  ) +
  
  labs(
    x = "Sample size n",
    y = "Hop diameter",
    title = "Growth of the hop diameter",
    subtitle = paste0(
      "Dashed curve: ",
      round(C_addario, 4),
      " n^(1/3)"
    )
  ) +
  theme_bw() +
  theme(
    aspect.ratio = 1,
    plot.title = element_text(face = "bold"),
    legend.position = "bottom"
  )
```

## Combining the Penrose and Addario-Berry rates

For a weighted tree,

$$
D_n \le H_n M_n,
$$

where $H_n$ is the hop diameter and $M_n$ is the largest edge weight.
This motivates combining the $n^{1/3}$ rate from @addarioberry2006diameter with the
Penrose rate $\sqrt{\log(n)/n}$:

$$
g(n)
=
n^{1/3}\sqrt{\frac{\log n}{n}}
=
n^{-1/6}\sqrt{\log n}.
$$


```{r}
comparison_summary <- combined_results |>
  group_by(n) |>
  summarise(
    mean_stringy_diameter = mean(stringy_diameter),
    .groups = "drop"
  ) |>
  mutate(
    combined_rate =
      n^(1/3) * sqrt(log(n) / n)
  )
```

Estimate one constant $C$ so that

$$
D_n \approx Cg(n).
$$

```{r}
multiplied_fit <- lm(
  mean_stringy_diameter ~ 0 + combined_rate,
  data = comparison_summary
)

C_multiplied <- unname(coef(multiplied_fit)[1])

C_multiplied
```

Create the smooth benchmark curve:

```{r}
multiplied_curve <- data.frame(
  n = seq(
    min(combined_results$n),
    max(combined_results$n),
    length.out = 500
  )
) |>
  mutate(
    theoretical_product =
      C_multiplied *
      n^(1/3) *
      sqrt(log(n) / n)
  )
```

Then plot only the simulations, their mean, and the combined rate:

```{r}
ggplot() +
  
  geom_point(
    data = combined_results,
    aes(
      x = n,
      y = stringy_diameter,
      colour = "Simulation values"
    ),
    alpha = 0.3,
    size = 0.5
  ) +
  
  geom_line(
    data = comparison_summary,
    aes(
      x = n,
      y = mean_stringy_diameter,
      colour = "Empirical mean"
    ),
    linewidth = 0.7
  ) +
  
  geom_point(
    data = comparison_summary,
    aes(
      x = n,
      y = mean_stringy_diameter,
      colour = "Empirical mean"
    ),
    size = 1.2
  ) +
  
  geom_line(
    data = multiplied_curve,
    aes(
      x = n,
      y = theoretical_product,
      colour = "Combined rate"
    ),
    linewidth = 1,
    linetype = "dashed"
  ) +
  
  scale_colour_manual(
    name = NULL,
    values = c(
      "Simulation values" = "black",
      "Empirical mean" = "royalblue",
      "Combined rate" = "violetred"
    )
  ) +
  
  labs(
    x = "Sample size n",
    y = "Weighted MST diameter",
    title = "Stringy diameter and combined growth rate",
    subtitle = paste0(
      "Dashed curve: ",
      round(C_multiplied, 4),
      " n^(-1/6) sqrt(log(n))"
    )
  ) +
  
  theme_bw() +
  theme(
    aspect.ratio = 1,
    plot.title = element_text(face = "bold"),
    legend.position = "bottom"
  )
```

## Exploring the ratio after normalizing the denominator

For bivariate uniform data, the total Euclidean MST length $L_n$, has a theoretically established $\sqrt{n}$ growth rate. More specifically, the result gives almost-sure convergence of $L_n/\sqrt{n}$ to a positive constant. We therefore use this known rate to remove the sample-size growth of the denominator while leaving the weighted MST diameter $D_n$ unchanged, since no corresponding theoretical growth rate is currently available for the numerator. For each individual simulation, we calculate

$$
R_n^*
=
\frac{D_n}{L_n/\sqrt{n}}.
$$

The purpose is to examine how the numerator behaves relative to the theoretically normalized denominator and whether the resulting empirical distribution shows any tendency to stabilize as $n$ increases. 

```{r}
uniform_files <- c(
  "stringy05_uniform_50_175.csv",
  "stringy05_uniform_200_500.csv",
  "stringy05_uniform_550_800.csv",
  "stringy05_uniform_850_1000.csv",
  "stringy05_uniform_1100_2000.csv",
  "stringy05_uniform_3000_5000.csv"
)

uniform_results <- map_dfr(
  uniform_files,
  ~ read_csv(.x, show_col_types = FALSE)
) |>
  arrange(n, simulation_run)


uniform_ratio_results <- uniform_results |>
  mutate(
    # Theoretical normalization of denominator
    normalized_mst_length = total_mst_length / sqrt(n),

    # Numerator remains same
    ratio_denominator_scaled =
      stringy_diameter / normalized_mst_length
  )

ratio_summary <- uniform_ratio_results |>
  group_by(n) |>
  summarise(
    mean_ratio = mean(
      ratio_denominator_scaled,
      na.rm = TRUE
    ),
    .groups = "drop"
  )

```

```{r}
# Normalize the denominator by sqrt(n)
length_scaled_results <- uniform_ratio_results |>
  mutate(
    length_over_sqrtn = total_mst_length / sqrt(n)
  )

length_scaled_summary <- length_scaled_results |>
  group_by(n) |>
  summarise(
    mean_length_sqrtn =
      mean(length_over_sqrtn, na.rm = TRUE),
    .groups = "drop"
  )

# Plot L_n / sqrt(n)
ggplot() +
  geom_point(
    data = length_scaled_results,
    aes(
      x = n,
      y = length_over_sqrtn,
      colour = "Simulation values"
    ),
    alpha = 0.3,
    size = 0.6
  ) +
  geom_line(
    data = length_scaled_summary,
    aes(
      x = n,
      y = mean_length_sqrtn,
      colour = "Empirical mean"
    ),
    linewidth = 0.7
  ) +
  geom_point(
    data = length_scaled_summary,
    aes(
      x = n,
      y = mean_length_sqrtn,
      colour = "Empirical mean"
    ),
    size = 1.2
  ) +
  scale_colour_manual(
    name = NULL,
    values = c(
      "Simulation values" = "black",
      "Empirical mean" = "steelblue"
    )
  ) +
  labs(
    x = "Sample size n",
    y = expression(L[n] / sqrt(n)),
    title = expression(
      "Total MST length normalized by " * sqrt(n)
    )
  ) +
  theme_bw() +
  theme(
    aspect.ratio = 1,
    plot.title = element_text(face = "bold"),
    legend.position = "bottom"
  )
```


## Plot the normalized ratio and the original ratio side by side


```{r}

# original Stringy05 ratio
uniform_ratio_results <- uniform_ratio_results |>
  mutate(
    original_ratio = stringy_diameter / total_mst_length
  )

original_ratio_summary <- uniform_ratio_results |>
  group_by(n) |>
  summarise(
    mean_original_ratio = mean(original_ratio, na.rm = TRUE),
    .groups = "drop"
  )

# --------------------------------------------------
# Left: ratio after normalizing denominator

p_scaled <- ggplot() +
  geom_point(
    data = uniform_ratio_results,
    aes(
      x = n,
      y = ratio_denominator_scaled,
      colour = "Simulation values"
    ),
    alpha = 0.3,
    size = 0.6
  ) +
  geom_line(
    data = ratio_summary,
    aes(
      x = n,
      y = mean_ratio,
      colour = "Empirical mean"
    ),
    linewidth = 0.7
  ) +
  geom_point(
    data = ratio_summary,
    aes(
      x = n,
      y = mean_ratio,
      colour = "Empirical mean"
    ),
    size = 1.2
  ) +
  scale_colour_manual(
    name = NULL,
    values = c(
      "Simulation values" = "black",
      "Empirical mean" = "steelblue"
    )
  ) +
  labs(
    x = "Sample size n",
    y = expression(
      D[n] / (L[n] / sqrt(n))
    ),
    title = "Ratio after normalizing denominator"
  ) +
  theme_bw() +
  theme(
    aspect.ratio = 1,
    plot.title = element_text(face = "bold"),
    legend.position = "bottom"
  )


# --------------------------------------------------
# Right: original Stringy05 ratio

p_original <- ggplot() +
  geom_point(
    data = uniform_ratio_results,
    aes(
      x = n,
      y = original_ratio,
      colour = "Simulation values"
    ),
    alpha = 0.3,
    size = 0.6
  ) +
  geom_line(
    data = original_ratio_summary,
    aes(
      x = n,
      y = mean_original_ratio,
      colour = "Empirical mean"
    ),
    linewidth = 0.7
  ) +
  geom_point(
    data = original_ratio_summary,
    aes(
      x = n,
      y = mean_original_ratio,
      colour = "Empirical mean"
    ),
    size = 1.2
  ) +
  scale_colour_manual(
    name = NULL,
    values = c(
      "Simulation values" = "black",
      "Empirical mean" = "steelblue"
    )
  ) +
  labs(
    x = "Sample size n",
    y = expression(D[n] / L[n]),
    title = "Original Stringy05 ratio"
  ) +
  theme_bw() +
  theme(
    aspect.ratio = 1,
    plot.title = element_text(face = "bold"),
    legend.position = "bottom"
  )


p_scaled + p_original +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```


> Next, we examine the weighted MST diameter $D_n$, which forms the numerator of `stringy05`, relative to the same $\sqrt{n}$ scale used for the denominator. Since no theoretical growth rate is currently available for $D_n$, dividing it by $\sqrt{n}$ is used only as an empirical comparison to determine whether the numerator grows on the same scale as the total MST length.


```{r}
# Normalize the numerator by sqrt(n)

uniform_ratio_results <- uniform_ratio_results |>
  mutate(
    normalized_diameter =
      stringy_diameter / sqrt(n)
  )

diameter_scaled_summary <- uniform_ratio_results |>
  group_by(n) |>
  summarise(
    mean_normalized_diameter =
      mean(normalized_diameter, na.rm = TRUE),
    .groups = "drop"
  )

```

### Normalized weighted diameter

```{r}
p_diameter_scaled <- ggplot() +
  geom_point(
    data = uniform_ratio_results,
    aes(
      x = n,
      y = normalized_diameter,
      colour = "Simulation values"
    ),
    alpha = 0.3,
    size = 0.6
  ) +
  geom_line(
    data = diameter_scaled_summary,
    aes(
      x = n,
      y = mean_normalized_diameter,
      colour = "Empirical mean"
    ),
    linewidth = 0.7
  ) +
  geom_point(
    data = diameter_scaled_summary,
    aes(
      x = n,
      y = mean_normalized_diameter,
      colour = "Empirical mean"
    ),
    size = 1.2
  ) +
  scale_colour_manual(
    name = NULL,
    values = c(
      "Simulation values" = "black",
      "Empirical mean" = "steelblue"
    )
  ) +
  labs(
    x = "Sample size n",
    y = expression(D[n] / sqrt(n)),
    title = "Weighted diameter normalized by sqrt(n)"
  ) +
  theme_bw() +
  theme(
    aspect.ratio = 1,
    plot.title = element_text(face = "bold"),
    legend.position = "bottom"
  )

p_diameter_scaled
```
> Since normalizing the weighted MST diameter by $\sqrt n$ produces a clear decreasing trend, the simulations suggest that the diameter grows more slowly than $\sqrt n$. We therefore compare several slower candidate functions of sample size. For each candidate rate $g(n)$, the weighted diameter is normalized as $D_n/g(n)$. A suitable empirical scaling should remove most of the sample-size trend, so that the normalized diameter becomes approximately stable for large $n$.


```{r}
library(tidyr)
# Compare candidate growth rates for the diameter
# --------------------------------------------------

diameter_rate_results <- uniform_ratio_results |>
  transmute(
    n,
    simulation_run,
    stringy_diameter,

    `sqrt(n)`      = stringy_diameter / sqrt(n),
    `n^(1/3)`      = stringy_diameter / n^(1/3),
    `n^(1/4)`      = stringy_diameter / n^(1/4),
    `n^(1/5)`      = stringy_diameter / n^(1/5),
    `log(n)`       = stringy_diameter / log(n),
    `sqrt(log(n))` = stringy_diameter / sqrt(log(n))
  ) |>
  pivot_longer(
    cols = -c(n, simulation_run, stringy_diameter),
    names_to = "normalization",
    values_to = "normalized_diameter"
  )

diameter_rate_results <- diameter_rate_results |>
  mutate(
    normalization = factor(
      normalization,
      levels = c(
        "sqrt(n)",
        "n^(1/3)",
        "n^(1/4)",
        "n^(1/5)",
        "log(n)",
        "sqrt(log(n))"
      )
    )
  )
```

Calculate the mean after normalizing each individual simulation:

```{r}
diameter_rate_summary <- diameter_rate_results |>
  group_by(n, normalization) |>
  summarise(
    mean_normalized_diameter =
      mean(normalized_diameter, na.rm = TRUE),
    .groups = "drop"
  )
```

### Plot all candidate normalizations

```{r}
ggplot() +
  geom_point(
    data = diameter_rate_results,
    aes(
      x = n,
      y = normalized_diameter
    ),
    alpha = 0.15,
    size = 0.45
  ) +
  geom_line(
    data = diameter_rate_summary,
    aes(
      x = n,
      y = mean_normalized_diameter
    ),
    colour = "steelblue",
    linewidth = 0.8
  ) +
  geom_point(
    data = diameter_rate_summary,
    aes(
      x = n,
      y = mean_normalized_diameter
    ),
    colour = "steelblue",
    size = 1.1
  ) +
  facet_wrap(
    ~ normalization,
    scales = "free_y",
    ncol = 3
  ) +
  labs(
    x = "Sample size n",
    y = expression(D[n] / g(n)),
    title = "Candidate normalizations for the weighted MST diameter",
    subtitle = "Blue line shows the empirical mean"
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(face = "bold"),
    strip.text = element_text(face = "bold")
  )
```



>  The normalized diameter continues to decrease under $\sqrt{n}$, $n^{1/3}$, and $n^{1/4}$, indicating that these functions increase faster than the empirical weighted-diameter growth. In contrast, normalization by $\sqrt{\log n}$ produces an increasing trend, suggesting that this rate is too slow. The $n^{1/5}$ and $\log n$ normalizations give the flattest curves, with $\log n$ appearing particularly stable at larger sample sizes. Since these comparisons are empirical, the next step is to quantify stability rather than selecting a rate only by visual inspection.


### 1. Compare slopes after $n=1000$

A good candidate should have a slope close to zero.

```{r}
rate_slopes <- diameter_rate_summary |>
  filter(n >= 1000) |>
  group_by(normalization) |>
  summarise(
    slope = coef(
      lm(mean_normalized_diameter ~ n)
    )[2],
    .groups = "drop"
  ) |>
  mutate(
    abs_slope = abs(slope)
  ) |>
  arrange(abs_slope)

rate_slopes
```

The smallest `abs_slope` ($log(n)$) is the flattest candidate.


### 2. Compare relative change over large $n$


```{r}
rate_stability <- diameter_rate_summary |>
  filter(n >= 1000) |>
  group_by(normalization) |>
  summarise(
    min_value = min(mean_normalized_diameter),
    max_value = max(mean_normalized_diameter),
    mean_value = mean(mean_normalized_diameter),

    relative_range =
      (max_value - min_value) / mean_value,

    .groups = "drop"
  ) |>
  arrange(relative_range)

rate_stability
```

> Among the candidate functions considered, $\log n$ provides the most stable empirical normalization over $n\ge1000$.





> The previous analysis suggested that $\log n$ provides the most stable empirical normalization for the weighted MST diameter $D_n$, while the total MST length $L_n$ has a theoretically established $\sqrt n$ scaling. We therefore normalize the numerator and denominator separately using these two rates. 


```{r}
# Normalize numerator by log(n)
# and denominator by sqrt(n)

normalized_ratio_results <- uniform_results |>
  mutate(
    diameter_over_logn =
      stringy_diameter / log(n),

    length_over_sqrtn =
      total_mst_length / sqrt(n),

    normalized_ratio =
      diameter_over_logn / length_over_sqrtn
  )
```

Summarise after calculating the ratio for each simulation:

```{r}
normalized_ratio_summary <- normalized_ratio_results |>
  group_by(n) |>
  summarise(
    mean_normalized_ratio =
      mean(normalized_ratio, na.rm = TRUE),

    median_normalized_ratio =
      median(normalized_ratio, na.rm = TRUE),

    .groups = "drop"
  )
```


```{r}
ggplot() +
  geom_point(
    data = normalized_ratio_results,
    aes(
      x = n,
      y = normalized_ratio,
      colour = "Simulation values"
    ),
    alpha = 0.3,
    size = 0.6
  ) +
  geom_line(
    data = normalized_ratio_summary,
    aes(
      x = n,
      y = mean_normalized_ratio,
      colour = "Empirical mean"
    ),
    linewidth = 0.8
  ) +
  geom_point(
    data = normalized_ratio_summary,
    aes(
      x = n,
      y = mean_normalized_ratio,
      colour = "Empirical mean"
    ),
    size = 1.2
  ) +
  scale_colour_manual(
    name = NULL,
    values = c(
      "Simulation values" = "black",
      "Empirical mean" = "steelblue"
    )
  ) +
  labs(
    x = "Sample size n",
    y = expression(
      (D[n] / log(n)) /
        (L[n] / sqrt(n))
    ),
    title = "Stringy05 after normalization",
    subtitle = expression(
      D[n] / log(n) ~ "and" ~ L[n] / sqrt(n)
    )
  ) +
  theme_bw() +
  theme(
    aspect.ratio = 1,
    plot.title = element_text(face = "bold"),
    legend.position = "bottom"
  )
```

> After applying the separate normalizations $D_n/\log n$ for the weighted diameter and $L_n/\sqrt n$ for the total MST length, the resulting ratio appears much more stable across sample size. This suggests that the empirical $\log n$ scaling for the numerator, together with the theoretical $\sqrt n$ scaling for the denominator, removes most of the observed sample-size trend in the ratio over the simulated range. However, the simulation values still show noticeable spread, so the next step is to examine whether the variance, upper quantiles, and overall distributional shape also stabilize.



## Distributional behaviour of the normalized ratio


The previous empirical comparison suggested that $\log n$ provides the most stable normalization for the weighted MST diameter among the candidate functions considered. However, stabilization of the empirical mean alone is not sufficient to conclude that the statistic has become sample-size invariant. The variance, upper tail, and overall shape of the distribution may still change with $n$. This is particularly important for `stringy05`, since the practical calibration of the index is based on the 95th percentile of its noise distribution rather than its mean. We therefore examine how the location, spread, upper quantiles, and distributional shape of the normalized `stringy05` ratio change with sample size.


### Summarise the distribution of the normalized ratio

```{r}
# before normalization
original_distribution_summary <- uniform_results |>
  mutate(
    original_ratio =
      stringy_diameter / total_mst_length
  ) |>
  group_by(n) |>
  summarise(
    mean = mean(original_ratio, na.rm = TRUE),
    q95 = quantile(original_ratio, 0.95, na.rm = TRUE),
    sd = sd(original_ratio, na.rm = TRUE),
    .groups = "drop"
  )

# after normalization
normalized_distribution_summary <- normalized_ratio_results |>
  group_by(n) |>
  summarise(
    mean = mean(normalized_ratio, na.rm = TRUE),
    median = median(normalized_ratio, na.rm = TRUE),

    variance = var(normalized_ratio, na.rm = TRUE),
    sd = sd(normalized_ratio, na.rm = TRUE),
    iqr = IQR(normalized_ratio, na.rm = TRUE),
    mad = mad(normalized_ratio, na.rm = TRUE),

    q90 = quantile(normalized_ratio, 0.90, na.rm = TRUE),
    q95 = quantile(normalized_ratio, 0.95, na.rm = TRUE),
    q99 = quantile(normalized_ratio, 0.99, na.rm = TRUE),

    .groups = "drop"
  )
```


For the proposed normalization to remove most of the sample-size dependence,
we would expect the mean, median, spread measures, and upper quantiles to become
approximately stable for sufficiently large $n$.


## Examine whether the spread stabilizes

```{r}

# Before normalization
p_sd_original <- ggplot(
  original_distribution_summary,
  aes(x = n, y = sd)
) +
  geom_line(
    colour = "steelblue",
    linewidth = 0.8
  ) +
  geom_point(
    colour = "steelblue"
  ) +
  labs(
    x = "Sample size n",
    y = "Standard deviation",
    title = "SD before normalization"
  ) +
  theme_bw() +
  theme(
    aspect.ratio = 1,
    plot.title = element_text(face = "bold")
  )


# After normalization
p_sd_normalized <- ggplot(
  normalized_distribution_summary,
  aes(x = n, y = sd)
) +
  geom_line(
    colour = "steelblue",
    linewidth = 0.8
  ) +
  geom_point(
    colour = "steelblue"
  ) +
  labs(
    x = "Sample size n",
    y = "Standard deviation",
    title = "SD after normalization"
  ) +
  theme_bw() +
  theme(
    aspect.ratio = 1,
    plot.title = element_text(face = "bold")
  )

p_sd_original + p_sd_normalized
```

The standard deviation is larger after normalization because the normalized statistic is on a different scale, so the two SD values are not directly comparable. What matters more is how the SD changes with sample size. Before normalization, the SD decreases strongly as $n$ increases, while after normalization it stays within a much narrower range. This suggests that the normalization reduces the sample-size dependence of the spread, although some variation across $n$ still remains.


## Compare the centre and the 95th percentile

```{r}
# Summary before normalization
original_distribution_summary <- uniform_results |>
  mutate(
    original_ratio = stringy_diameter / total_mst_length
  ) |>
  group_by(n) |>
  summarise(
    mean_original = mean(original_ratio, na.rm = TRUE),
    q95_original = quantile(original_ratio, 0.95, na.rm = TRUE),
    .groups = "drop"
  )

# Join with normalized summary
comparison_summary <- original_distribution_summary |>
  left_join(
    normalized_distribution_summary |>
      select(
        n,
        mean_normalized = mean,
        q95_normalized = q95
      ),
    by = "n"
  )

# Plot all four together
ggplot(comparison_summary, aes(x = n)) +
  geom_line(
    aes(
      y = mean_original,
      colour = "Mean before normalization"
    ),
    linewidth = 0.8
  ) +
  geom_point(
    aes(
      y = mean_original,
      colour = "Mean before normalization"
    )
  ) +
  geom_line(
    aes(
      y = mean_normalized,
      colour = "Mean after normalization"
    ),
    linewidth = 0.8
  ) +
  geom_point(
    aes(
      y = mean_normalized,
      colour = "Mean after normalization"
    )
  ) +
  geom_line(
    aes(
      y = q95_original,
      colour = "95th percentile before normalization"
    ),
    linewidth = 0.8
  ) +
  geom_point(
    aes(
      y = q95_original,
      colour = "95th percentile before normalization"
    )
  ) +
  geom_line(
    aes(
      y = q95_normalized,
      colour = "95th percentile after normalization"
    ),
    linewidth = 0.8
  ) +
  geom_point(
    aes(
      y = q95_normalized,
      colour = "95th percentile after normalization"
    )
  ) +
  scale_colour_manual(
    name = NULL,
    values = c(
      "Mean before normalization" = "steelblue",
      "Mean after normalization" = "darkorange",
      "95th percentile before normalization" = "violetred",
      "95th percentile after normalization" = "darkgreen"
    )
  ) +
  labs(
    x = "Sample size n",
    y = "Ratio value",
    title = "Mean and 95th percentile before and after normalization"
  ) +
  theme_bw() +
  theme(
    legend.position = "bottom"
  )
```


> The normalized ratio appears more stable across sample sizes; however, the 95th percentile of the normalized statistic can exceed 1. This is an important limitation because scagnostic indices are intended to lie between 0 and 1. The normalization therefore removes the natural upper bound of the original `stringy05` definition. Using this normalized ratio directly as a new scagnostic would require an additional transformation or rescaling to return the values to $[0,1]$, which adds another layer of complexity to the definition.










# Appendix

## Alternative way

### Scaling limits of the Euclidean minimum spanning tree

A potentially useful theoretical connection for understanding the sample-size
behaviour of the Stringy index is provided by
@aizenman1999scaling.


Aizenman et al. study continuum scaling limits of several random spanning-tree
models in two dimensions. Of particular relevance here is their **Euclidean
minimal spanning tree (EST)**. The vertices of the EST are generated from a
homogeneous Poisson point process in $\mathbb{R}^2$ with intensity
$\delta^{-2}$. Every pair of points can be connected, the weight of an edge is
its ordinary Euclidean distance, and the Euclidean MST is the spanning tree
that minimises the sum of these Euclidean edge lengths
[@aizenman1999scaling].


Thus, if $X_i$ and $X_j$ denote two points, the edge weight is

$$
w_{ij} = \lVert X_i-X_j\rVert_2.
$$

This setup is important because the edge weights are generated by Euclidean
geometry rather than being independent random edge weights. It is therefore
substantially closer to the MST used in the Stringy index than random
complete-graph MST models based on independent Uniform$(0,1)$ edge weights.

Aizenman et al. investigate what happens as

$$
\delta \rightarrow 0,
$$
so that the Poisson point process becomes increasingly dense. This can be viewed of studying an increasingly large number of
points in a fixed two-dimensional region.

### Main findings

Aizenman et al. establish tightness of the distributions of these spanning
trees and show that subsequential continuum scaling limits exist
[@aizenman1999scaling]. Their main geometric finding is that the branches of
the limiting Euclidean MST are random curves with non-trivial fractal
geometry.


In particular, the Hausdorff dimension of the limiting branches is bounded by

$$
1 < d_{\min}
\leq
\dim_H(C)
\leq
d_{\max} < 2,
$$

for non-random constants $d_{\min}$ and $d_{\max}$
[@aizenman1999scaling].

The fact that the dimension is strictly greater than one means that these
branches are substantially rougher than ordinary smooth curves. Aizenman
et al. prove, more strongly, that the limiting branches are
**nonrectifiable** [@aizenman1999scaling; @aizenman1999holder].

A rectifiable curve has finite ordinary arc length. For example, a straight
line, circle, or sufficiently smooth curve has a finite length that is
approached as the curve is measured at increasingly fine resolutions.
A nonrectifiable curve does not have finite ordinary arc length. Increasingly
fine resolution reveals additional geometric detail, so the polygonal length
does not settle to a finite limiting value
[@aizenman1999holder; @mandelbrot1967coast].

