---
title: "Exploring the Growth Rate of the stringy05 Index"
date: 2026-07-30
categories: [scagnostics, minimum spanning tree, stringy05, asymptotic analysis, R]
description: "A theoretical and empirical exploration of the stringy05 scagnostic index."
license: "All Rights Reserved"
copyright: "Copyright © 2026 Tina Rashid Jafari. All rights reserved."
format:
  html:
    toc: true
    toc-depth: 3
execute:
  echo: true
code-fold: true
warning: FALSE
message: FALSE
---

## Aim

The goal of this blog is to understand the asymptotic behavior of the `stringy05` index, defined as

$$
\text{stringy05} =
\frac{{diameter}(T_n)}{{length}(T_n)}.
$$

Here $T_n$ is the Euclidean minimum spanning tree (MST) built on $n$ random bivariate observations.

The denominator is the total Euclidean length of the MST,

$$
L_n
= {length}(T_n) = \sum_{e\in T_n}|e|.
$$

The numerator is the weighted graph diameter of the MST,

$$
D_n
= {diameter}(T_n)
= \max_{u,v\in T_n} d_{T_n}(u,v),
$$
where $d_{T_n}(u,v)$ is the length of the unique path in the tree from $u$ to $v$, using Euclidean edge lengths as weights.

Thus,

$$
R_n
=
\text{stringy05}(T_n)
=
\frac{D_n}{L_n}.
$$

The main question is whether

$$
R_n \to 0,
$$

or whether

$$
R_n \to \rho > 0.
$$

Empirically, previous simulations suggested that the ratio does not rapidly decay toward zero. Around sample sizes near $n=900$, the estimated 95th percentile of the noise distribution appeared to stabilize near $0.1$. This motivates a more detailed investigation of the numerator and denominator separately.


## Denominator: theoretical growth rate of total MST length

The denominator $L_n$ is the part of the problem for which we do have strong theoretical guidance.

Steele studies Euclidean minimal spanning trees with power-weighted edges. In his notation, for points $x_1,\dots,x_n$, the MST functional is

$$
M(x_1,\dots,x_n)
=
\min_T \sum_{e\in T}\psi(|e|),
$$

where the minimum is over spanning trees $T$, and $\psi$ is an edge-weighting function [Steele, 1988].

For the ordinary Euclidean MST length, we take

$$
\psi(t)=t.
$$

Then

$$
M(X_1,\dots,X_n)
=
\sum_{e\in T_n}|e|
=
L_n.
$$

So Steele's $M_n$ is exactly the denominator of the stringy05 index.

## Steele's growth theorem

Steele considers weight functions satisfying

$$
\psi(t)\sim t^\alpha
\quad\text{as } t\to 0,
$$

with

$$
0<\alpha<d.
$$


His main theorem states that, for iid observations in $\mathbb R^d$ from a compactly supported distribution, if $f$ is the density of the absolutely continuous part of the distribution, then

$$
n^{-(d-\alpha)/d}
M(X_1,\dots,X_n)
\to
c(\alpha,d)
\int f(x)^{(d-\alpha)/d}\,dx
$$

almost surely.


In the ordinary MST length case,

$$
\alpha=1.
$$

For bivariate data,

$$
d=2.
$$

Therefore,

$$
n^{-1/2}L_n
\to
c(1,2)
\int f(x)^{1/2}\,dx.
$$

Equivalently,

$$
L_n
\sim
C_f\sqrt n,
$$

where

$$
C_f
=
c(1,2)
\int f(x)^{1/2}\,dx.
$$

Thus, for two-dimensional data, the denominator has theoretical growth rate

$$
\boxed{
L_n= O(\sqrt n).
}
$$

Steele first proves the MST growth-rate result in the uniform setting, where the sample points are distributed in the unit cube. Steele then extends the result beyond the uniform case by adding a density correction. For a more general distribution with density $f$, the limiting constant is modified by a factor of the form

$$
\int f(x)^{(d-\alpha)/d}\,dx.
$$

In the ordinary bivariate MST-length case, this becomes

$$
\int f(x)^{1/2}\,dx.
$$

This extension is important for my project because the simulations are based on bivariate noise distributions rather than only uniform samples.

However, there is one important technical caveat. Steele's theorem is stated under compact-support assumptions. A standard bivariate normal distribution has unbounded support, so it falls outside the direct assumptions of the theorem. Therefore, for Gaussian samples, I use Steele's$\sqrt n$ scaling as a theoretically motivated benchmark for the denominator, not as a theorem that applies directly without further work. 


## Numerator: what is known about the MST diameter?

The numerator is

$$
D_n
=
\max_{u,v}d_{T_n}(u,v).
$$

This is the weighted longest shortest path in the MST. It is a global path functional: it depends on how many edges lie along the longest path and how their lengths accumulate.

This is different from the total MST length $L_n$, which is a sum over all edges. It is also different from the longest single edge of the MST.

At this point, the exact theoretical growth rate of $D_n$ is not given by the Steele MST length theorem.


## Penrose's longest-edge theorem and its limitation

Penrose studies the longest edge of the random Euclidean MST. Let

$$
E_n
=
\max_{e\in T_n}|e|.
$$

For uniform random points in the unit square, Penrose shows that the longest MST edge has the same asymptotic extreme-value behavior as the longest nearest-neighbor edge. In dimension $2$, the result implies that

$$
n\pi E_n^2-\log n
$$

converges in distribution to a double-exponential limiting law [Penrose, 1997].


Penrose also shows that very long MST edges usually behave like nearest-neighbor edges. In other words, the longest edges are typically attached to leaf vertices, or endpoints, of the MST.

This matters for the stringy05 index because the MST diameter is not just the longest single edge. The diameter is the length of a whole path through the tree. So Penrose's result suggests that the longest individual edges are often at the outside of the tree, rather than forming the main interior path that determines the MST diameter.


The longest edge scale is therefore approximately

$$
E_n
\asymp
\sqrt{\frac{\log n}{n}}.
$$
However, this is not the same as the stringy05 numerator.

The stringy05 numerator is

$$
D_n =
\max_{u,v} d_{T_n}(u,v),
$$

which is the length of an entire path. This path may contain many MST edges. Therefore, Penrose's theorem does not give the growth rate of $D_n$. It only gives the scale of the largest individual edge.


## A simple example: longest edge versus MST diameter

Let's imagine the MST is a path with five vertices:

$$
A \;-\; B \;-\; C \;-\; D \;-\; E
$$

and the edge weights are

$$
|AB|=0.10,\quad |BC|=0.12,\quad |CD|=0.09,\quad |DE|=0.11.
$$

Penrose's longest-edge quantity looks only for the single largest edge in the MST:

$$
E_n=\max_{e\in T_n}|e|=0.12.
$$

So in this example, Penrose's quantity is the edge $BC$.

But the stringy05 numerator is the weighted MST diameter. It looks for the longest path in the tree and adds all edge weights along that path. Here the longest path is from $A$) to $E$:

$$
D_n = |AB|+|BC|+|CD|+|DE| =
0.10+0.12+0.09+0.11 =
0.42.
$$

So in this example,

$$
\text{Penrose longest edge}=0.12,
$$

while

$$
\text{stringy05 diameter}=0.42.
$$


```{r}
library(cassowaryr)
library(igraph)
library(ggplot2)
library(patchwork)

set.seed(1050)

n <- 200
u <- runif(n, -1, 1)
polynomial <- poly(u, degree = 2)

x <- polynomial[, 1] + rnorm(n, sd = 0.03)
y <- polynomial[, 2] + rnorm(n, sd = 0.03)

ggplot(data.frame(x, y), aes(x, y)) +
  geom_point(size = 1.6) +
  coord_equal() +
  theme_bw() +
  labs(
    x = "Polynomial degree 1",
    y = "Polynomial degree 2"
  ) +
  theme(aspect.ratio = 1,
        axis.ticks = element_blank(),
        axis.text = element_blank())
```

## Construct the minimum spanning tree

```{r}
sc <- scree(
  x,
  y,
  out.rm = FALSE,
  binner = NULL
)

mst <- cassowaryr:::gen_mst(
  sc$del,
  sc$weights
)

mst_weights <- E(mst)$weight
```

## Calculate the two diameter definitions

```{r}
# Stringy: longest weighted path through the MST
diameter_path <- get_diameter(
  mst,
  directed = FALSE,
  weights = mst_weights
)

weighted_diameter <- diameter(
  mst,
  directed = FALSE,
  weights = mst_weights
)

# Penrose: longest individual MST edge
longest_edge <- E(mst)[which.max(mst_weights)]
penrose_diameter <- longest_edge$weight

# Stringy05
stringy05 <- weighted_diameter / sum(mst_weights)

cat("Penrose longest edge:", round(penrose_diameter, 5), "\n")
cat("Weighted MST diameter:", round(weighted_diameter, 5), "\n")
cat("Stringy05:", round(stringy05, 5), "\n")
```

## Draw the MST and highlight the two diameters

```{r}
# Coordinates of the Stringy diameter path
diameter_xy <- as.data.frame(
  sc$del$x[as.integer(diameter_path), ]
)

names(diameter_xy) <- c("x", "y")

# Coordinates of the longest individual edge
penrose_vertices <- ends(
  mst,
  longest_edge,
  names = FALSE
)

penrose_xy <- as.data.frame(
  sc$del$x[penrose_vertices, ]
)

names(penrose_xy) <- c("x", "y")

# Base MST plot supplied by cassowaryr
mst_plot <- draw_mst(mst, sc) +
  coord_equal() +
  theme_bw()

p_stringy <- mst_plot +
  geom_path(
    data = diameter_xy,
    aes(x, y),
    colour = "#0072B2",
    linewidth = 1.3
  ) +
  geom_point(
    data = diameter_xy[c(1, nrow(diameter_xy)), ],
    aes(x, y),
    colour = "#0072B2",
    size = 3
  ) +
  labs(
    title = "Stringy weighted diameter",
    subtitle = paste(
      "Weighted diameter =",
      round(weighted_diameter, 4)
    )
  )

p_penrose <- mst_plot +
  geom_segment(
    data = penrose_xy,
    aes(
      x = x[1],
      y = y[1],
      xend = x[2],
      yend = y[2]
    ),
    colour = "#D55E00",
    linewidth = 1.5
  ) +
  geom_point(
    data = penrose_xy,
    aes(x, y),
    colour = "#D55E00",
    size = 3
  ) +
  labs(
    title = "Penrose longest MST edge",
    subtitle = paste(
      "Longest edge =",
      round(penrose_diameter, 4)
    )
  )

p_stringy + p_penrose +
  plot_annotation(
    subtitle = paste(
      "Stringy05 =",
      round(stringy05, 4)
    )
  )
```


This shows why Penrose's result is useful only as a weak lower bound. The longest single edge is always less than or equal to the weighted MST diameter, but it is not the same quantity:

$$
E_n\le D_n.
$$


## Upper and lower bound for the MST diameter

Even though Penrose's longest-edge theorem does not solve the diameter problem, it provides a weak lower bound.

For every weighted tree,

$$
\boxed{
E_n\le D_n\le L_n.
}
$$

The first inequality holds because the longest edge is itself the path between its endpoints in the tree. The second inequality holds because any path in the tree uses only a subset of the edges of the tree. 


Combining Penrose's longest-edge scale with Steele's total-length scale gives

$$
\sqrt{\frac{\log n}{n}}
\lesssim
D_n
\le
O(\sqrt n).
$$

Dividing by $L_n$ gives the weak theoretical ratio bound

$$
\frac{\sqrt{\log n}}{n}
\lesssim
R_n
\le
1.
$$

This lower bound is mathematically valid but very weak. For example, at $n=1000$,

$$
\frac{\sqrt{\log(1000)}}{1000}
\approx
0.00263.
$$

This is much smaller than the empirical values observed in simulation, which are closer to $0.2$. Therefore, the observed stringy05 behavior cannot be explained by the longest single MST edge alone. The diameter appears to be driven by the accumulation of many edges along a long path.



## Next step: study numerator and denominator separately

The next part of this analysis will study

$$
L_n=\operatorname{length}(T_n)
$$

and

$$
D_n=\operatorname{diameter}(T_n)
$$

separately.

The plan is:

1. Generate bivariate noise samples for a grid of sample sizes $n$.
2. Record the total MST length $L_n$ .
3. Record the weighted MST diameter $D_n$.
4. Record the ratio

$$
R_n=\frac{D_n}{L_n}.
$$

For the denominator, Steele's theorem suggests checking whether

$$
\frac{L_n}{\sqrt n}
$$

stabilizes.

For the numerator, the key empirical question is whether

$$
\frac{D_n}{\sqrt n}
$$

also stabilizes.

If both normalized quantities stabilize, then the ratio should stabilize near a positive constant:

$$
R_n
=
\frac{D_n}{L_n}
\approx
\frac{C_D\sqrt n}{C_L\sqrt n}
=
\frac{C_D}{C_L}.
$$

This would support the empirical observation that stringy05 converges to a positive number, possibly near $0.1$, rather than to zero.



## Empirical check of the denominator growth rate

To check the theoretical growth rate of the denominator, I simulated independent bivariate observations from the uniform distribution on the unit square, which matches the setting used in Steele's theoretical result. For each sample size, I repeated the simulation 100 times (just for testing), constructed the Euclidean minimum spanning tree, and recorded its total edge length,

$$
L_n = \sum_{e\in T_n}|e|.
$$

The simulation results were saved in separate files for different ranges of sample sizes and are combined below.

```{r}
library(readr)
library(dplyr)
library(purrr)
library(ggplot2)

csv_files <- c(
  "stringy05_uniform_50_175.csv",
  "stringy05_uniform_200_500.csv",
  "stringy05_uniform_550_800.csv",
  "stringy05_uniform_850_1000.csv",
  "stringy05_uniform_1100_2000.csv",
  "stringy05_uniform_3000_5000.csv"
)

combined_results <- map_dfr(
  csv_files,
  ~ read_csv(.x, show_col_types = FALSE)
) |>
  arrange(n, simulation_run)
```

Steele's result implies that, for bivariate uniform data, the total MST length grows at rate

$$
L_n \sim C\sqrt n,
$$

for some positive constant $C$.

To compare the simulations with this theoretical rate, I first calculate the empirical mean MST length for each sample size. I then estimate the multiplicative constant $C$ by fitting the relationship

$$
L_n \approx C\sqrt n
$$

through the origin.

```{r}
# Calculate the mean for each sample size
length_summary <- combined_results |>
  group_by(n) |>
  summarise(
    mean_length = mean(total_mst_length),
    .groups = "drop"
  )

# Estimate C in L_n = C * sqrt(n)
length_fit <- lm(
  mean_length ~ 0 + I(sqrt(n)),
  data = length_summary
)

C_hat <- unname(coef(length_fit)[1])

C_hat

# Theoretical curve based on the fitted constant
theoretical_curve <- data.frame(
  n = seq(
    min(combined_results$n),
    max(combined_results$n),
    length.out = 500
  )
) |>
  mutate(
    theoretical_length = C_hat * sqrt(n)
  )
```

The following plot shows all individual simulation values of the total MST length. The blue line gives the empirical mean at each sample size, while the pink dashed line shows the fitted theoretical function

$$
\widehat C\sqrt n.
$$

If the denominator follows the theoretical $\sqrt n$ growth rate, the empirical mean should have approximately the same shape as this dashed curve.

```{r}
ggplot() +
  geom_point(
    data = combined_results,
    aes(
      x = n,
      y = total_mst_length,
      colour = "Simulation values"
    ),
    alpha = 0.7,
    size = 0.5
  ) +
  geom_line(
    data = length_summary,
    aes(
      x = n,
      y = mean_length,
      colour = "Empirical mean"
    ),
    linewidth = 0.5
  ) +
  geom_point(
    data = length_summary,
    aes(
      x = n,
      y = mean_length,
      colour = "Empirical mean"
    ),
    size = 1
  ) +
  geom_line(
    data = theoretical_curve,
    aes(
      x = n,
      y = theoretical_length,
      colour = "Theoretical sqrt(n)"
    ),
    linewidth = 1,
    linetype = "dashed"
  ) +
  scale_colour_manual(
    name = NULL,
    values = c(
      "Simulation values" = "black",
      "Empirical mean" = "royalblue",
      "Theoretical sqrt(n)" = "violetred"
    )
  ) +
  labs(
    x = "Sample size n",
    y = "Total MST length",
    title = "Growth of total Euclidean MST length",
    subtitle = paste0(
      "Dashed curve: ",
      round(C_hat, 4),
      " sqrt(n)"
    )
  ) +
  theme_bw() +
  theme(
    aspect.ratio = 1,
    plot.title = element_text(face = "bold"),
    legend.position = "bottom"
  )
```

A second way to check the growth rate is to remove the proposed $\sqrt n$ scaling. If

$$
L_n \sim C\sqrt n,
$$

then

$$
\frac{L_n}{\sqrt n}\to C.
$$

Therefore, if the theoretical rate is appropriate, the normalized MST length should become increasingly stable as the sample size grows.

The plot below shows the normalized value $L_n/\sqrt n$ for every simulation. The blue line gives the empirical mean for each sample size, and the pink dashed line shows the same finite-sample estimate $\widehat C$ obtained above.

```{r}
# Normalize total MST length by sqrt(n)
combined_results <- combined_results |>
  mutate(
    length_over_sqrtn = total_mst_length / sqrt(n)
  )

# Mean normalized length for each sample size
length_scaled_summary <- combined_results |>
  group_by(n) |>
  summarise(
    mean_length_over_sqrtn = mean(length_over_sqrtn),
    .groups = "drop"
  )

ggplot() +
  geom_point(
    data = combined_results,
    aes(
      x = n,
      y = length_over_sqrtn,
      colour = "Simulation values"
    ),
    alpha = 0.7,
    size = 0.5
  ) +
  geom_line(
    data = length_scaled_summary,
    aes(
      x = n,
      y = mean_length_over_sqrtn,
      colour = "Empirical mean"
    ),
    linewidth = 0.5
  ) +
  geom_point(
    data = length_scaled_summary,
    aes(
      x = n,
      y = mean_length_over_sqrtn,
      colour = "Empirical mean"
    ),
    size = 1
  ) +
  geom_hline(
    aes(
      yintercept = C_hat,
      colour = "Estimated C"
    ),
    linewidth = 1,
    linetype = "dashed"
  ) +
  scale_colour_manual(
    name = NULL,
    values = c(
      "Simulation values" = "black",
      "Empirical mean" = "royalblue",
      "Estimated C" = "violetred"
    )
  ) +
  labs(
    x = "Sample size n",
    y = expression(L[n] / sqrt(n)),
    title = expression("Convergence of " * L[n] / sqrt(n)),
    subtitle = paste0(
      "Dashed horizontal line: C = ",
      round(C_hat, 4)
    )
  ) +
  theme_bw() +
  theme(
    aspect.ratio = 1,
    plot.title = element_text(face = "bold"),
    legend.position = "bottom"
  )
```



## Empirical check of the longest MST edge

The theoretical result for the numerator is not directly available for the weighted MST diameter used by `stringy05`. However, Penrose studies the longest **single edge** of the Euclidean minimum spanning tree. Let

$$
E_n = \max_{e \in T_n} |e|.
$$

For independent uniform observations on the unit square, the longest MST edge has asymptotic scale

$$
E_n \asymp \sqrt{\frac{\log n}{n}}.
$$

This is different from the actual `stringy05` numerator, which is the weighted length of an entire path through the MST. Therefore, in the simulations I use the stored `penrose_diameter` variable to check Penrose's longest-edge result separately.

For each sample size, the simulation was repeated 100 times. I first calculate the empirical mean longest-edge length for each value of $n$.

```{r}
penrose_summary <- combined_results |>
  group_by(n) |>
  summarise(
    mean_penrose = mean(penrose_diameter),
    .groups = "drop"
  )
```

Penrose's result suggests a relationship of the form

$$
E_n \approx A
\sqrt{\frac{\log n}{n}},
$$

where $A$ is a positive constant.

As with the denominator analysis, I estimate the constant $A$ from the simulation values.

```{r}
penrose_fit <- lm(
  mean_penrose ~ 0 + I(sqrt(log(n) / n)),
  data = penrose_summary
)

A_hat <- unname(coef(penrose_fit)[1])

A_hat
```

The estimated value $\widehat A$ is then used to construct the fitted theoretical curve

$$
\widehat A
\sqrt{\frac{\log n}{n}}.
$$

```{r}
penrose_theoretical_curve <- data.frame(
  n = seq(
    min(combined_results$n),
    max(combined_results$n),
    length.out = 500
  )
) |>
  mutate(
    theoretical_penrose =
      A_hat * sqrt(log(n) / n)
  )
```

The following plot shows all individual simulation values of the longest MST edge. The blue line shows the empirical mean for each sample size, while the pink dashed curve shows the fitted theoretical function

$$
\widehat A
\sqrt{\frac{\log n}{n}}.
$$

If the simulations are consistent with Penrose's result, the empirical longest-edge values should decrease with approximately the same shape as the theoretical curve.

```{r}
ggplot() +
  geom_point(
    data = combined_results,
    aes(
      x = n,
      y = penrose_diameter,
      colour = "Simulation values"
    ),
    alpha = 0.7,
    size = 0.5
  ) +
  geom_line(
    data = penrose_summary,
    aes(
      x = n,
      y = mean_penrose,
      colour = "Empirical mean"
    ),
    linewidth = 0.5
  ) +
  geom_point(
    data = penrose_summary,
    aes(
      x = n,
      y = mean_penrose,
      colour = "Empirical mean"
    ),
    size = 1
  ) +
  geom_line(
    data = penrose_theoretical_curve,
    aes(
      x = n,
      y = theoretical_penrose,
      colour = "Theoretical sqrt(log(n)/n)"
    ),
    linewidth = 1,
    linetype = "dashed"
  ) +
  scale_colour_manual(
    name = NULL,
    values = c(
      "Simulation values" = "black",
      "Empirical mean" = "royalblue",
      "Theoretical sqrt(log(n)/n)" = "violetred"
    )
  ) +
  labs(
    x = "Sample size n",
    y = "Longest MST edge",
    title = "Scaling of the longest Euclidean MST edge",
    subtitle = paste0(
      "Dashed curve: ",
      round(A_hat, 4),
      " sqrt(log(n)/n)"
    )
  ) +
  theme_bw() +
  theme(
    aspect.ratio = 1,
    plot.title = element_text(face = "bold"),
    legend.position = "bottom"
  )
```



## Does the stringy05 numerator follow the Penrose rate?

The previous analysis verified the theoretical result for the longest single MST edge. However, the numerator of `stringy05` is not the longest individual edge. It is the weighted MST diameter, which is the total Euclidean length of the longest path through the MST.


The next empirical question is whether the weighted diameter nevertheless follows a similar scaling,

$$
D_n
\approx
B\sqrt{\frac{\log n}{n}},
$$

for some constant $B$.

```{r}
stringy_diameter_summary <- combined_results |>
  group_by(n) |>
  summarise(
    mean_stringy_diameter = mean(stringy_diameter),
    .groups = "drop"
  )
```

To compare the weighted diameter with the Penrose functional form, I fit

$$
D_n
\approx
B\sqrt{\frac{\log n}{n}}
$$

 The purpose is to test whether the **shape and rate** of the weighted diameter are consistent with the Penrose rate.

```{r}
stringy_diameter_fit <- lm(
  stringy_diameter ~ 0 + I(sqrt(log(n) / n)),
  data = combined_results
)

B_hat <- unname(coef(stringy_diameter_fit)[1])

B_hat
```

```{r}
stringy_theoretical_curve <- data.frame(
  n = seq(
    min(combined_results$n),
    max(combined_results$n),
    length.out = 500
  )
) |>
  mutate(
    theoretical_stringy_diameter =
      B_hat * sqrt(log(n) / n)
  )
```

The following plot compares all simulated weighted MST diameters with the Penrose's theoretical boundary.

```{r}
ggplot() +
  geom_point(
    data = combined_results,
    aes(
      x = n,
      y = stringy_diameter,
      colour = "Simulation values"
    ),
    alpha = 0.7,
    size = 0.5
  ) +
  geom_line(
    data = stringy_diameter_summary,
    aes(
      x = n,
      y = mean_stringy_diameter,
      colour = "Empirical mean"
    ),
    linewidth = 0.5
  ) +
  geom_point(
    data = stringy_diameter_summary,
    aes(
      x = n,
      y = mean_stringy_diameter,
      colour = "Empirical mean"
    ),
    size = 1
  ) +
  geom_line(
    data = stringy_theoretical_curve,
    aes(
      x = n,
      y = theoretical_stringy_diameter,
      colour = "Penrose-type rate"
    ),
    linewidth = 1,
    linetype = "dashed"
  ) +
  scale_colour_manual(
    name = NULL,
    values = c(
      "Simulation values" = "black",
      "Empirical mean" = "royalblue",
      "Penrose-type rate" = "violetred"
    )
  ) +
  labs(
    x = "Sample size n",
    y = "Weighted MST diameter",
    subtitle = paste0(
      "Dashed curve: ",
      round(B_hat, 4),
      " sqrt(log(n)/n)"
    )
  ) +
  theme_bw() +
  theme(
    aspect.ratio = 1,
    plot.title = element_text(face = "bold"),
    legend.position = "bottom"
  )
```


## Addario-Berry, Broutin and Reed: an $n^{1/3}$ MST-diameter benchmark

Addario-Berry, Broutin and Reed study a different random minimum-spanning-tree model. They begin with the complete graph $K_n$, assign independent and identically distributed continuous weights to all edges, and construct the minimum-weight spanning tree. Their main result is

$$
\mathbb{E}\{\operatorname{diam}(T_n)\}
=
\Theta(n^{1/3}).
$$

There are two important differences from my setting:

1. **Randomness of the edge weights.** In Addario-Berry et al., the edge weights of $K_n$ are i.i.d. In my Euclidean MST, the edge weights are distances between random points in $\mathbb{R}^2$, so the pairwise edge weights are geometrically constrained and are not independent.
2. **Definition of diameter.** Their diameter is the ordinary graph diameter, based on graph distance along the tree. My `stringy05` numerator is a **weighted Euclidean diameter**, where the Euclidean edge lengths are summed along the longest weighted path.

To see whether the empirical weighted diameter has anything resembling the same growth shape, I fit the benchmark

$$
D_n
\approx
C_{\mathrm{ABR}} n^{1/3},
$$

to the empirical weighted-diameter means.

```{r}

abr_fit <- lm(
  mean_stringy_diameter ~ 0 + I(n^(1/3)),
  data = stringy_diameter_summary
)

C_ABR_hat <- unname(coef(abr_fit)[1])


abr_theoretical_curve <- data.frame(
  n = seq(
    min(combined_results$n),
    max(combined_results$n),
    length.out = 500
  )
) |>
  mutate(
    abr_n13_benchmark = C_ABR_hat * n^(1/3)
  )
```

The following plot shows the same simulated weighted MST diameters used in the previous section, but now compares them with the fitted $n^{1/3}$ benchmark from Addario-Berry et al.

```{r}
ggplot() +
  geom_point(
    data = combined_results,
    aes(
      x = n,
      y = stringy_diameter,
      colour = "Simulation values"
    ),
    alpha = 0.7,
    size = 0.5
  ) +
  geom_line(
    data = stringy_diameter_summary,
    aes(
      x = n,
      y = mean_stringy_diameter,
      colour = "Empirical mean"
    ),
    linewidth = 0.5
  ) +
  geom_point(
    data = stringy_diameter_summary,
    aes(
      x = n,
      y = mean_stringy_diameter,
      colour = "Empirical mean"
    ),
    size = 1
  ) +
  geom_line(
    data = abr_theoretical_curve,
    aes(
      x = n,
      y = abr_n13_benchmark,
      colour = "Addario-Berry n^(1/3) benchmark"
    ),
    linewidth = 1,
    linetype = "dashed"
  ) +
  scale_colour_manual(
    name = NULL,
    values = c(
      "Simulation values" = "black",
      "Empirical mean" = "royalblue",
      "Addario-Berry n^(1/3) benchmark" = "violetred"
    )
  ) +
  labs(
    x = "Sample size n",
    y = "Weighted MST diameter",
    subtitle = paste0(
      "Dashed curve: ",
      round(C_ABR_hat, 4),
      " n^(1/3)"
    )
  ) +
  theme_bw() +
  theme(
    aspect.ratio = 1,
    plot.title = element_text(face = "bold"),
    legend.position = "bottom"
  )
```



## References

Steele, J. M. (1988). *Growth Rates of Euclidean Minimal Spanning Trees with Power Weighted Edges*. The Annals of Probability, 16(4), 1767--1787.

Penrose, M. D. (1997). *The Longest Edge of the Random Minimal Spanning Tree*. The Annals of Applied Probability, 7(2), 340--361.

Addario-Berry, L., Broutin, N., & Reed, B. (2006). *The Diameter of the Minimum Spanning Tree of a Complete Graph*. Discrete Mathematics & Theoretical Computer Science, DMTCS Proceedings vol. AG, 237--248. https://doi.org/10.46298/dmtcs.3513